---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# datavyur

<!-- badges: start -->
<!-- badges: end -->

The goal of datavyur is to provide set of functions to handle import/export between R and the qualitative coding software Datavyu.

## Installation

You can install the released version of datavyur from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("datavyur")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("iamamutt/datavyur")
```

## Example

This is a basic example which shows you how to import Datavyu data into R. First, load the package.

```{r load_pkg}
# Load the datavyur library to use the functions below
library(datavyur)
```

Next, point to a directory containing `.csv` data that has been previously exported using the script `datavyu2csv.rb` (see the section "Exporting Datavyu `.opf` files" below). This example uses the internal package data directory for demonstration purposes.

```{r example_data}
ex_data_dir <- datavyur_internal_data()
```

```{r, echo=FALSE, warning=FALSE}
dv_info <- datavyu_search(folder=ex_data_dir)
```

The folder `ex_data_dir` originally contained `r length(unique(dv_info$file))` `.opf` files that have been split into `r length(unique(dv_info$local))` separate `.csv` files, one `.csv` file for each Datavyu file (`.opf`) and column combination.

Names of `.csv` files after using the Ruby script `datavyu2csv.rb`:

```{r, echo=FALSE}
list.files(ex_data_dir, pattern="\\.csv$")
```

Names of the original Datavyu `.opf` files:

```{r, echo=FALSE}
unique(dv_info$file)
```
Names of the Datavyu columns found inside the set of exported `.csv` files:

```{r, echo=FALSE}
unique(dv_info$column)
```

This information can be obtained by using the function `datavyu_search`.

```{r, eval=FALSE}
datavyu_search(folder=ex_data_dir)
```

Use the `import_datavyu` function to import Datavyu columns into R as separate `data.frames`.
Since there are multiple Datavyu columns that need to be imported from `ex_data_dir`, set `as_list=TRUE` to import a `data.frame` for each Datavyu column.

```{r, warning=FALSE}
dv_columns <- import_datavyu(folder=ex_data_dir, as_list=TRUE)
```

You can access each `data.frame` like so:

```{r}
child_data <- dv_columns$childhands
```

The first few rows are shown below.

```{r, echo=FALSE}
knitr::kable(head(child_data))
```

To learn more, see the Getting Started vignette.
