---
title: "Data Manipulation"
author: "Joseph M. Burling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results='hide', message=FALSE, warning = FALSE}
knitr::opts_chunk$set(eval = TRUE, comment = ">", 
                      message = FALSE, 
                      warning = FALSE)
library(datavyur)
library(dplyr)
```

# Importing the example data

First we'll import the example data that comes with the `datavyur` package. Assume that there were originally 3 `.opf` files (`dyad1.opf`, `dyad2.opf`, and `dyad3.opf`) and the script `datavyu2csv.rb` was used on these files, which then created a `.csv` for each file and Datavyu column. We'll import these data and align them by file and time information by converting timestamps to frame numbers and then merging all columns using the single function `temporal_align`. Note that the Datavyu path isn't set. We didn't set `options(datavyur.folder)` because by default, the folder is set to the example data used in this package.

```{r}
# load the packages we are going to be using, must be installed first
library(datavyur)
library(dplyr)

# align data from chilhands and parenthands columns by file and frame number
mydata <- temporal_align()
```

We'll also clean up the data to make sure invalid codes and timestamps are marked as `NA`.

```{r}
# create list of valid codes
codes_to_check  <- list(
        childhands.hand = c("left", "right", "both"),
        childhands.look = c(1, 0),
        parenthands.hand = c("left", "right", "both"),
        parenthands.look = c(1, 0)
    )

# create list of timestamp columns
timestamps_to_check <- list(
    child = c(on="childhands.onset", off="childhands.offset"),
    parent = c(on="parenthands.onset", off="parenthands.offset")
)

# check and overwrite existing data
mydata <- check_codes(codes_to_check, mydata)$data
mydata <- check_timestamps(timestamps_to_check, mydata)$data
```

Showing the first 6 and last 6 rows of the data:

```{r, echo=FALSE}
knitr::kable(rbind(head(mydata), tail(mydata)))
```

# Data manipulation using `dplyr`

The R package `dplyr` is a good add-on package to start with when learning basic data manipulation techniques. There are many different approaches toward data manipulation, such as learning the standard R language with `data.frame` objects and the standard `apply()` family of functions in R, or using a more powerful data structure called `data.table` (an alternative to `dplyr`), but these require going into foundations of the R language and learning more complicated syntax than just going with the `dplyr` approach. Since the goal is to understand data manipulation for absolute beginners and not computer science persons, we'll stick with `dplyr` for now.

A more detailed introduction to `dplyr` can be found by viewing the intro vignette by typing the following code in the console.

```{r, eval=FALSE}
vignette("introduction", "dplyr")
```

# Understanding the `dplyr` "verbs"

These are the most common set of operations you'll perform on the data. Combinations of these in your scripts can get you pretty far if you know how to use them creatively. For more documentation on each of these, prefix the function name with a `?` into the console like so.

```{r, eval=FALSE}
# get help on dplyr::filter
?filter
```

You might see multiple functions called `filter`, but click on the one corresponding to the `dplyr` library.

*****

# `select`, `rename`

These two functions deal with columns in a data set. The `select()` function allows you to choose specific columns that you are interested in, while the `rename()` function sets new names for existing columns.

Let us select and only work with the `parenthands` columns and omit any columns regarding `childhands` for now. 

Here are several ways of achieving the same thing:

```{r, eval=FALSE}
# 1. from mydata, select all id and parenthands.x columns
parent <- select(mydata, file, frame_number, parenthands.ordinal, parenthands.onset, parenthands.offset, parenthands.hand, parenthands.look)

# 2. use ranges if the column order is known
parent <- select(mydata, file, frame_number, parenthands.ordinal:parenthands.look)

# 3. by omitting the childhands columns using the - sign
parent <- select(mydata, -(childhands.ordinal:childhands.look))
```

```{r, echo=FALSE}
parent <- select(mydata, -(childhands.ordinal:childhands.look))
```

The `rename()` function just overwrites existing columns with new names. This might be helpful if some columns are too long and it makes it easier to call it something shorter, or if the information in a column changes and you want to reflect these changes with a new name.

Here we'll rename the current column called `frame_number` to something shorter, like `frame`. After specifying the data to work with, the new name is on the left of the `=` sign, and the old name on the right. The new data with the new column name is saved over the old data (`parent`).

```{r}
# in parent data, rename frame_number to frame, overwrite same data
parent <- rename(parent, frame = frame_number)
```

A sample of the data, notice how the same events happen for multiple frames of time. The number of rows for these data is proportional to the duration from onset to offset.

```{r, echo=FALSE}
knitr::kable(head(filter(parent, !is.na(parenthands.hand))))
```

*****

# `distinct`

The `distinct()` function allows you to remove any duplicated rows from the data. You specify the data, then specify which columns to check for unique rows. If you don't specify columns, it defaults to using all the columns in the data.

From the `parent` data we made in the previous step, we can remove the column `frame`. The rest of the data will then have duplicates of the same information across rows (see table above). This is because we originally imported data using the `temporal_align()` function from `datavyur`. If importing only a single column using `import_column()`, we wouldn't have duplicate information. Use the `distinct()` function to remove these duplicates.

```{r}
# get rid of frame column from parent data
parent <- select(parent, -(frame))

# obtain distinct rows, checking all columns in the parent data
parent <- distinct(parent)
```

You'll now see much less rows than before using `distinct()`.

```{r, echo=FALSE}
knitr::kable(head(filter(parent, file == "dyad1", parenthands.ordinal >= 2)))
```

*****

# `mutate`, `transmute`

The `mutate()` function operates across columns of data, and will change or alter data in an existing column with new data, or create a new column from manipulating existing data. We can create a new column that calculates the duration of timestamps from already existing onset and offset data. We'll call this new column `parenthands.duration`. Notice how this new column is a function of two different columns in the parent data.

### Mutated data

```{r}
# make new column called parenthands.duration
parent <- mutate(parent, parenthands.duration = parenthands.offset - parenthands.onset)
```

```{r, echo=FALSE}
knitr::kable(head(parent))
```

### Transmuted data

```{r, eval=FALSE}
# transmute does the same but keeps only the columns that are specified
parent <- transmute(parent, file, parenthands.duration = parenthands.offset - parenthands.onset)
```

```{r, echo=FALSE}
knitr::kable(head(transmute(parent, file, parenthands.duration = parenthands.offset - parenthands.onset)))
```

*****

# `filter`, `slice`

Whereas `select()` deals with columns in your data, the functions `filter()` and `slice()` deal with rows of data. Basically, you're trying to filter out rows that meet some criteria you specify by using logical operators such as **AND** and **OR**.

Now let us analyze the data to see if there are moments when the child is looking at both of the parent's hands. We'll filter the data used previously and save this filtered data as a new data object called `both_hand_look`. 

```{r}
# each condition separated by a comman
# filter(data, something, AND something else, AND this too)
both_hand_look <- filter(parent, parenthands.hand == "both", parenthands.look == 1)
```

Notice the double `=` sign written as `==`. This means "is equivalent to", and should not be confused with a single equals sign. Also, the commas in the `filter()` function between condition serve as the logical operator `&`.

This is the entire data set that met those conditions.

```{r, echo=FALSE}
knitr::kable(both_hand_look)
```

We can also check for a logical **OR** by using the `|` character instead of commas. Here we'll check for `left` or `right` parent hands, but leave out `both`.

```{r}
# hands == left OR right, AND look == 1
lr_hand_look <- filter(parent, parenthands.hand == "left" | parenthands.hand == "right", parenthands.look == 1)
```

And here are the data that meet the above conditions. Notice that the `parenthands.hand` column contains `left` and `right` values.

```{r, echo=FALSE}
knitr::kable(lr_hand_look)
```

If you want to subset your data by row numbers instead of by conditions, use the `slice()` function instead.


```{r, eval=FALSE}
# select rows 1 through 10 from the parent data set
slice(parent, 1:10)

# select rows 1,5,10,100
slice(parent, c(1, 5, 10, 100))
```

*****

# `arrange`

The `arrange()` function sorts your data from smallest to largest by the columns you specify.

```{r}
# arrange data from shortest to longest duration
lr_hand_look <- arrange(lr_hand_look, parenthands.duration)
```

To arrange in descending order (largest to smallest), use the `desc()` function inside the `arrange()` function like so.

```{r}
# arrange data from shortest to longest duration
lr_hand_look <- arrange(lr_hand_look, desc(parenthands.duration))
```

How the data look by duration in descending order.

```{r, echo=FALSE}
knitr::kable(lr_hand_look)
```

*****

# `summarise`

Now we've reached the point where we can start creating summaries of our data and calculating some basic statistics. The `summarise()` function takes multiple rows of data and collapses them into a single value. You should use any function that will return a single value. Here we are using the `mean()` function, to take several observations and return a single value of central tendency. In the same operation, we'll also convert the mean from milliseconds to seconds by dividing by 1000. This new value will be called `avg_sec`, and will be store in the data object `seconds_stat`. 

```{r}
# mean of time spent fixated on left or right hands in seconds across all dyads
seconds_stat <- summarise(lr_hand_look, avg_sec = mean(parenthands.duration) / 1000)
```

```{r, echo=FALSE}
seconds_stat
```


# Applying verbs over groups of data

A lot of the power of `dplyr` comes from using the above set of functions on chunks of data instead of applying them over the entire data set. For example, above, we took the mean duration collapsing across all dyads. If we wanted to do this operation for each dyad separately, we would use the function `group_by` along with the `summarise` function.

First we'll create a new data set that sets the groups we want. Here, we'll set the data into groups obtained from the `file` column, which is the dyad id. We'll go back to the original `parent` data set that has all values for `parenthands.hand`.


```{r}
# group parent data by file, save as new data set
parent_grouped <- group_by(parent, file)
```

Once the data has been grouped and saved, we can now take the grouped data and apply the same `summarise()`, operation as before. This time, we'll set the optional argument `na.rm` in the `mean()` function to `TRUE` so that we remove `NA`s before trying to find the mean, otherwise the output will be `NA`.

```{r}
# same operation, but on grouped data, and removing any NAs
summarise(parent_grouped, avg_sec = mean(parenthands.duration, na.rm = TRUE) / 1000)
```

See how instead of one value we now get 3 values, 1 for each dyad? The `group_by` function can be very powerful, and group by more than one column. Here we group by `file` and `hand` code, then run the same operation using `summarise()`.

```{r}
# remove NAs from parenthands.hand first
# use is.na to check NA values within parenthands.hand, return the opposite using ! (NOT)
parent <- filter(parent, !is.na(parenthands.hand))

# group parent data by file and hand, save as new data set
parent_grouped2 <- group_by(parent, file, parenthands.hand)

summarise(parent_grouped2, avg_sec = mean(parenthands.duration, na.rm = TRUE) / 1000)
```

Now we have an average duration for each dyad and for each value within `parenthands.hand`. If you wish to save these results, make sure to assign the output to a variable name to store it.

*****

# Chaining operations together

The techniques shown above require saving a lot of intermediate data sets, such as having to save grouped data then performing some summary operation on the grouped data. You can be much more concise by chaining several of the `dplyr` functions together using the special set of characters `%>%` after each command.

For example, let's repeat the steps we took starting way back from the original data called `mydata`, but use the chaining method. The first line in the chain starts off with the name of the data, then each function performed in order after that. Since we specified the data at the top, we don't have to keep specifying the data over and over again when using each of the "verb" actions.

```{r}
final_result <-                                                             # 0
    mydata %>%                                                              # 1
    select(file, childhands.ordinal:childhands.look) %>%                    # 2
    distinct() %>%                                                          # 3
    mutate(childhands.duration = childhands.offset - childhands.onset) %>%  # 4
    filter(childhands.hand == "both", childhands.look == 1) %>%             # 5
    group_by(file) %>%                                                      # 6
    summarise(avg_dur = mean(childhands.duration, na.rm = TRUE))            # 7
```

Here's a breakdown of what we did for each line, making sure that the end of the line contains the chaining command `%>%`, except for the last command which will send the output to `final_result`.

### Chain steps

0. Save the final output from the last line to the object `final_result`
1. Specify the data we're going to work with (`mydata`), and will be used throughout the chain
2. Select the columns `file`, and all `childhands.x` columns
3. Omit duplicate rows of data from the result in the previous step
4. Make a new column called `childhands.duration`, which is the total time for each cell calculated from the onset and offset columns
5. Filter columns only relating to `both` hands in view **AND** only during fixation moments (`x.look == 1`)
6. Group the we data generated thus far by `file` (which is the dyad id)
7. Find the average duration for each group specified from above, removing any `NA`'s when using the `mean()` function

The `final_result` output received from the chain of commands looks like the table below. There is no `dyad3` because for the `childhands` data, there were no valid observations returned for this dyad after executing step 5. 

```{r, echo=FALSE}
final_result
```

Now we have some summarized data. From here we can do some analyses or plotting within R, or we can export the summary data to another format to import into a different program.

```{r, eval=FALSE}
# export final_result data to a file called child_data_summary in your current working directory
write.csv(final_result, "child_data_summary.csv")
```

