---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datavyur)
```


## Datavyu data structure

It is important to understand the parts of one or more Datavyu files to understand how to use this R package. The structure is basically organized like so:

- datavyu files
  - column names
    - cells
      - standard fields <value>
      - custom fields <code>

For example, you may have several Datavyu files with multiple "columns" in each file, and in each column exists the standard `onset`, `offset`, and `ordinal` fields and their values, along with custom fields that hold their "codes."

```
|datavyu_file1.opf
|- column_name1
|  |- cell
|     |- onset <ts>
|     |- offset <ts>
|     |- ordinal <int>
|     |- custom_code_id1 <code>
|     |- custom_code_id2 <code>
|  |- cell
|     |- onset <ts>
|     |- offset <ts>
|     |- ordinal <int>
|     |- custom_code_id1 <code>
|     |- custom_code_id2 <code>
|  |- ...
|- column_name2
|  |- cell
|     |- onset <ts>
|     |- offset <ts>
|     |- ordinal <int>
|     |- custom_code_id1 <code>
|     |- custom_code_id2 <code>
|  |- ...
|datavyu_file2.opf
|- column_name1
|  |- ...
|- column_name2
|  |- ...
|...
```


## Example


This is a basic example which shows you how to import Datavyu data into R. First, load the package.

```{r load_pkg}
# Load the datavyur library to use the functions below
library(datavyur)
```

Next, point to a directory containing `.csv` data that has been previously exported using the script `datavyu2csv.rb` (see the section "Exporting Datavyu `.opf` files" below). This example uses the internal package data directory for demonstration purposes.

```{r example_data}
ex_data_dir <- datavyur_internal_data()
```

```{r, echo=FALSE, warning=FALSE}
dv_info <- datavyu_search(folder=ex_data_dir)
```

The folder `ex_data_dir` originally contained `r length(unique(dv_info$file))` `.opf` files that have been split into `r length(unique(dv_info$local))` separate `.csv` files, one `.csv` file for each Datavyu file (`.opf`) and column combination.

Names of `.csv` files after using the Ruby script `datavyu2csv.rb`:

```{r, echo=FALSE}
list.files(ex_data_dir, pattern="\\.csv$")
```

Names of the original Datavyu `.opf` files:

```{r, echo=FALSE}
unique(dv_info$file)
```
Names of the Datavyu columns found inside the set of exported `.csv` files:

```{r, echo=FALSE}
unique(dv_info$column)
```

This information can be obtained by using the function `datavyu_search`.

```{r, eval=FALSE}
datavyu_search(folder=ex_data_dir)
```

Use the `import_datavyu` function to import Datavyu columns into R as separate `data.frames`.
Since there are multiple Datavyu columns that need to be imported from `ex_data_dir`, set `as_list=TRUE` to import a `data.frame` for each Datavyu column.

```{r, warning=FALSE}
dv_columns <- import_datavyu(folder=ex_data_dir, as_list=TRUE)
```

You can access each `data.frame` like so:

```{r}
child_data <- dv_columns$childhands
```

The first few rows are shown below.

```{r, echo=FALSE}
knitr::kable(head(child_data))
```

## Horizontal alignment of Datavyu data

```{r, warning=FALSE}
main_list <- import_datavyu_to_list()
```

### Aligning data by timestamp

To align data found in the folder `ex_data_dir` use the function `temporal_align`.

```{r, warning=FALSE}
time_aligned1 <- temporal_align(folder=ex_data_dir, fps=30, columns="childhands")
```

```{r, echo=FALSE}
knitr::kable(head(time_aligned1))
```

### Aligning by ordinal value

```{r, warning=FALSE}
ord_aligned1 <- ordinal_align(folder=ex_data_dir, columns="parenthands")
```

```{r, echo=FALSE}
knitr::kable(head(ord_aligned1))
```

## Vertical alignment of Datavyu data

To stack several Datavyu columns' data vertically, use the function `vert_merge_datavyu_list`. This will convert all codes to characters so that they're able to exist in a single column.

```{r}
vert_data <- vert_merge_datavyu_list(main_list)
```

```{r, echo=FALSE}
knitr::kable(rbind(head(vert_data), tail(vert_data)))
```
